High geomagnetic field intensity in the late Mesoproterozoic recorded by Midcontinent Rift anorthosite xenoliths

### 2021-09-29 manuscript/Manuscript.tex

% need to export the figure with tight boundaries
\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\small{Summary plot of cooling rate corrected specimen absolute paleointensity results (square and diamond symbols) from this study and their weighted averages at site (grey bar) and locality level (orange bar). }}
\label{fig:PINT_cooling_corrected}
\end{figure}

@@

From the thermal history model we can also estimate the duration which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully block in their characteristic natural remanence component ($\sim$500\textdegree C; \citeA{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster, at about $1.3\times10^{-1}$ $^\circ$C s$^{-1}$ if we assume it takes 10 minutes in lab condition to cool through the same temperature interval. As a result, this significant cooling rate difference leads to a prediction of 35.5\% overestimate of true ancient field \cite{Halgedahl1980a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate corrected specimen paleointensity estimates together with specimen- and site-level weighted means are shown in Figure \ref{fig:PINT_cooling_corrected}. Taking all accepted anorthosite paleointensity results, they have a site mean absolute paleointensity estimates of 47.90 $\pm$ 9.32 $\mu$T (Figure \ref{fig:PINT_cooling_corrected}).

\subsection*{Late Mesoproterozoic geomagnetic field and the Precambrian observation records}
Figure \ref{fig:PINT_plot} shows all accepted anorthosite and diabase specimen paleointensity results and their site-level means with one standard deviation envelopes. An anorthosite site mean paleointensity weighted by site standard deviations is plotted as a dashed orange line, and the standard deviation envelope as dashed green lines. The anorthosites are also grouped by their geographical vicinity \ref{fig:PINT_plot}. The site-level results are shown in Table \ref{tab:PINT_result}. Anorthosite sites AX10, 11, 12, 13, 16 yielded very high within-site coherence (Figure \ref{fig:Arai_plots, fig:PINT_plot}), which suggest high-confidence site-level paleointensity estimates. Clearly there are two populations of paleointensity results: a group of high estimates from anorthosite site 11 and 13 which have very high within-site consistency and are in agreement with the results from BRD site 2; and another group of low estimates from anorthosite site 6, 8, 10, 12, and 16, which also have tight agreement across sites.

>>>>

% need to export the figure with tight boundaries
\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{Paleointensity_plot_cooling_corrected.pdf}
\centering
\caption{\small{Summary plot of cooling rate corrected individual specimen absolute paleointensity results (square and diamond symbols) from this study and their averages at site (grey bar) and locality level (weighted mean by site-level standard deviations; orange bar). The inset plot shows the graph of predicted paleointensity overestimate due to slow cooling of the intrusive Beaver River diabase relative to the cooling rate in laboratory following the model of \citeA{Halgedahl1980a}. Because the majority of the anorthosites have unblocking temperatures between 500\textdegree C and 580\textdegree C, we estimate that the slow cooling during natural remanence acquisition could have resulted in a maximum of 36\% overestimate. Thus a correction factor of 0.74 is applied at specimen level for the summary plot.  }}
\label{fig:PINT_cooling_corrected}
\end{figure}

@@

From the thermal history model we can also estimate the duration which the diabase and anorthosite cooled from the Curie temperature of magnetite ($\sim$580\textdegree C) to the time when they fully block in their characteristic natural remanence component ($\sim$500\textdegree C; \citeA{Zhang2021b}). We find the cooling time to be on the order of $\sim$1 kyr, which corresponds to a cooling rate of $1.6\times10^{-9}$ $^\circ$C s$^{-1}$. In contrast, the lab cooling rate is much faster, at about $1.3\times10^{-1}$ $^\circ$C s$^{-1}$ if we assume it takes 1000s in lab condition to cool through the same temperature interval. As a result, this significant cooling rate difference leads to a prediction of $\sim$36\% overestimate of true ancient field \cite{Halgedahl1980a}. We therefore correct our paleointensity results by a factor of 0.74. The cooling rate corrected specimen paleointensity estimates together with specimen- and site-level weighted means are shown in Figure \ref{fig:PINT_cooling_corrected}. Taking all accepted anorthosite paleointensity results, they have a site mean absolute paleointensity estimates of 47.90 $\pm$ 9.32 $\mu$T (Figure \ref{fig:PINT_cooling_corrected}).

\subsection*{Late Mesoproterozoic geomagnetic field and the Precambrian observation records}
The anorthosites are also grouped by their geographical vicinity \ref{fig:PINT_plot}. The site-level results are shown in Table \ref{tab:PINT_result}. Anorthosite sites AX10, 11, 12, 13, 16 yielded very high within-site coherence (Figure \ref{fig:Arai_plots, fig:PINT_plot}), which suggest high-confidence site-level paleointensity estimates. Clearly there are two populations of paleointensity results: a group of high estimates from anorthosite site 11 and 13 which have very high within-site consistency and are in agreement with the results from BRD site 2; and another group of low estimates from anorthosite site 6, 8, 10, 12, and 16, which also have tight agreement across sites.

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{PINT_BiCEP.pdf}
\centering
\caption{\footnotesize{Violin plot of paleointensity estimates using the bias corrected estimation of paleointensity (BiCEP) method \cite{Cych2021a}. }}
\label{fig:BiCEP}
\end{figure}

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{MCR_VADM.pdf}
\centering
\caption{\footnotesize{Compilation of calculated virtual axial dipole moment (VADM) values from previous studies of the Midcontinent Rift rocks \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} and from the anorthosites that pass the selection criteria in this study. The median VADM value of all MCR rocks is higher than the median value of the past 300 myr. }}
\label{fig:MCR_VADM}
\end{figure}
